---
title: "Gapminder Modern Dataset (1952–2022)"
subtitle: "Exploring global trends in population, life expectancy, GDP, and CO₂"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    fig-width: 9
    fig-height: 5.5
    df-print: kable
execute:
  warning: false
  message: false
---

## Overview

This report explores the **Gapminder Modern** dataset, which combines four indicators for up to 195 countries across 15 five-year snapshots from 1952 to 2022:

| Indicator | Description |
|-----------|-------------|
| `lifeExp` | Life expectancy at birth (years) |
| `pop` | Total population |
| `gdpPercap` | GDP per capita (inflation-adjusted USD) |
| `co2Percap` | CO₂ consumption per capita (tonnes) |

```{r}
#| label: setup
library(tidyverse)
library(scales)

gap <- read_csv("gapminder_modern_1952_2022.csv") |>
  filter(!is.na(continent))   # drop the 1 unmatched row

continent_colours <- c(
  Africa   = "#E69F00",
  Americas = "#56B4E9",
  Asia     = "#009E73",
  Europe   = "#CC79A7",
  Oceania  = "#F0E442"
)
```

## Dataset at a Glance

```{r}
#| label: tbl-overview
gap |>
  summarise(
    Countries  = n_distinct(country),
    Continents = n_distinct(continent),
    Years      = n_distinct(year),
    `Year range` = paste(min(year), "–", max(year)),
    Rows       = n(),
    `Missing lifeExp (%)` = round(mean(is.na(lifeExp)) * 100, 1),
    `Missing GDP (%)` = round(mean(is.na(gdpPercap)) * 100, 1),
    `Missing CO₂ (%)` = round(mean(is.na(co2Percap)) * 100, 1)
  )
```

```{r}
#| label: tbl-continent-counts
gap |>
  distinct(country, continent) |>
  count(continent, name = "Countries") |>
  arrange(desc(Countries))
```

---

## Life Expectancy

### Global Trend

Life expectancy has risen dramatically in every continent. The shaded ribbon shows the min–max country range; the line shows the population-weighted mean.

```{r}
#| label: fig-lifeexp-trend
#| fig-cap: "Population-weighted mean life expectancy by continent, 1952–2022"

gap |>
  filter(!is.na(lifeExp), !is.na(pop)) |>
  summarise(
    mean_le  = weighted.mean(lifeExp, pop),
    min_le   = min(lifeExp),
    max_le   = max(lifeExp),
    .by = c(continent, year)
  ) |>
  ggplot(aes(year, mean_le, colour = continent, fill = continent)) +
  geom_ribbon(aes(ymin = min_le, ymax = max_le), alpha = 0.12, colour = NA) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = continent_colours) +
  scale_fill_manual(values = continent_colours) +
  scale_x_continuous(breaks = seq(1952, 2022, 10)) +
  labs(x = NULL, y = "Life expectancy (years)",
       colour = NULL, fill = NULL)
```

### Distribution in 2022

```{r}
#| label: fig-lifeexp-2022
#| fig-cap: "Distribution of life expectancy by continent in 2022"

gap |>
  filter(year == 2022, !is.na(lifeExp)) |>
  ggplot(aes(lifeExp, fct_reorder(continent, lifeExp, median),
             colour = continent, fill = continent)) +
  ggridges::geom_density_ridges(alpha = 0.4, scale = 0.9) +
  scale_colour_manual(values = continent_colours) +
  scale_fill_manual(values = continent_colours) +
  labs(x = "Life expectancy (years)", y = NULL) +
  theme(legend.position = "none")
```

### Top & Bottom 10 Countries (2022)

```{r}
#| label: tbl-lifeexp-extremes
le_2022 <- gap |>
  filter(year == 2022, !is.na(lifeExp)) |>
  select(country, continent, lifeExp) |>
  arrange(desc(lifeExp))

bind_rows(
  slice_head(le_2022, n = 10) |> mutate(rank = "Top 10"),
  slice_tail(le_2022, n = 10) |> mutate(rank = "Bottom 10")
) |>
  select(rank, country, continent, `Life expectancy` = lifeExp)
```

---

## GDP per Capita

### Median GDP Trend by Continent

```{r}
#| label: fig-gdp-trend
#| fig-cap: "Median GDP per capita by continent, 1952–2022 (log scale)"

gap |>
  filter(!is.na(gdpPercap)) |>
  summarise(median_gdp = median(gdpPercap), .by = c(continent, year)) |>
  ggplot(aes(year, median_gdp, colour = continent)) +
  geom_line(linewidth = 1) +
  scale_y_log10(labels = scales::dollar_format()) +
  scale_colour_manual(values = continent_colours) +
  scale_x_continuous(breaks = seq(1952, 2022, 10)) +
  labs(x = NULL, y = "Median GDP per capita (USD, log scale)", colour = NULL)
```

### GDP Distribution in 2022

```{r}
#| label: fig-gdp-violin
#| fig-cap: "GDP per capita distribution by continent in 2022 (log scale)"

gap |>
  filter(year == 2022, !is.na(gdpPercap)) |>
  ggplot(aes(fct_reorder(continent, gdpPercap, median),
             gdpPercap, fill = continent)) +
  geom_violin(alpha = 0.6, draw_quantiles = 0.5) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.5) +
  scale_y_log10(labels = scales::dollar_format()) +
  scale_fill_manual(values = continent_colours) +
  labs(x = NULL, y = "GDP per capita (USD, log scale)") +
  theme(legend.position = "none")
```

---

## CO₂ per Capita

### Trend Over Time

```{r}
#| label: fig-co2-trend
#| fig-cap: "Population-weighted mean CO₂ per capita by continent, 1952–2022"

gap |>
  filter(!is.na(co2Percap), !is.na(pop)) |>
  summarise(
    mean_co2 = weighted.mean(co2Percap, pop),
    .by = c(continent, year)
  ) |>
  ggplot(aes(year, mean_co2, colour = continent)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = continent_colours) +
  scale_x_continuous(breaks = seq(1952, 2022, 10)) +
  labs(x = NULL, y = "CO₂ per capita (tonnes)", colour = NULL)
```

### Top Emitters in 2022

```{r}
#| label: tbl-co2-top
gap |>
  filter(year == 2022, !is.na(co2Percap)) |>
  slice_max(co2Percap, n = 15) |>
  select(Country = country, Continent = continent,
         `CO₂ per capita (t)` = co2Percap,
         `Population` = pop) |>
  mutate(`Population` = scales::number(Population, big.mark = ",", accuracy = 1))
```

---

## The Gapminder View: Wealth vs. Health

A bubble chart replicating the classic Gapminder "wealth vs. health" view for selected years.

```{r}
#| label: fig-bubble
#| fig-cap: "GDP per capita vs. life expectancy; bubble size = population"
#| fig-height: 7

gap |>
  filter(year %in% c(1952, 1977, 2002, 2022),
         !is.na(gdpPercap), !is.na(lifeExp), !is.na(pop)) |>
  ggplot(aes(gdpPercap, lifeExp,
             size = pop, colour = continent)) +
  geom_point(alpha = 0.55) +
  scale_x_log10(labels = scales::dollar_format()) +
  scale_size_area(max_size = 18,
                  labels = scales::label_number(big.mark = ",")) +
  scale_colour_manual(values = continent_colours) +
  facet_wrap(~year, ncol = 2) +
  labs(x = "GDP per capita (USD, log scale)",
       y = "Life expectancy (years)",
       colour = NULL, size = "Population") +
  theme(legend.position = "bottom")
```

---

## CO₂ vs. GDP

Wealthier countries tend to emit more CO₂, but the relationship has shifted over time.

```{r}
#| label: fig-co2-gdp
#| fig-cap: "CO₂ per capita vs. GDP per capita for 2022"

gap |>
  filter(year == 2022, !is.na(gdpPercap), !is.na(co2Percap)) |>
  ggplot(aes(gdpPercap, co2Percap,
             size = pop, colour = continent,
             label = country)) +
  geom_point(alpha = 0.6) +
  ggrepel::geom_text_repel(
    data = \(d) slice_max(d, co2Percap, n = 8),
    size = 3, max.overlaps = 20
  ) +
  scale_x_log10(labels = scales::dollar_format()) +
  scale_size_area(max_size = 16, guide = "none") +
  scale_colour_manual(values = continent_colours) +
  labs(x = "GDP per capita (USD, log scale)",
       y = "CO₂ per capita (tonnes)",
       colour = NULL)
```

---

## Summary Statistics (2022)

```{r}
#| label: tbl-summary-2022
gap |>
  filter(year == 2022) |>
  summarise(
    across(c(lifeExp, gdpPercap, co2Percap),
           list(median = \(x) median(x, na.rm = TRUE),
                mean   = \(x) mean(x, na.rm = TRUE),
                sd     = \(x) sd(x, na.rm = TRUE)),
           .names = "{.col}__{.fn}"),
    .by = continent
  ) |>
  pivot_longer(-continent, names_to = c("Indicator", "Stat"),
               names_sep = "__") |>
  pivot_wider(names_from = Stat, values_from = value) |>
  arrange(Indicator, continent) |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

---

## Data Notes

- Source: [Gapminder](https://www.gapminder.org/data/) (indicators: `lex`, `pop`, `gdp_pcap`, `co2_pcap_cons`).
- Years: every 5 years from 1952 to 2022 (15 snapshots).
- One unmatched ISO3 code with no country name was excluded (~15 rows).
- Small states (Liechtenstein, Monaco, San Marino) have partial data for some indicators.
- CO₂ values represent *consumption-based* emissions per capita.
